Poker Chips, Earthquakes, and Computation

### 2014-09-26 draft0.tex
changed up to lemma 3

A restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $GE^N(f) = GE^{N+m}(f)$ and $LE^{M}(f) = LE^{M+m}(f)$

\subsection{A Potential Function}
In order to show that if the state of a finite board cannot cycle under local or global earthquakes we define a potential function that maps board states to the nonnegative integers and show that applying either kind of earthquake to a board increases the value of the function. 
\begin{lemma}
If $GE(f) \neq f$, then $GE^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
Let $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
we are taking a sum of the number of chips on each square with a weight of the taxi cab distance from the square to the origin.
Note that this is well-defined since the boards contain a finite number of chips. We will show that

\[ GE(f) \neq f \Rightarrow \Phi(GE(f)) > \Phi(f) \]
That is if global earthquake changes the state of the board, then $\Phi(f)$ increases. Therefore if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or in other words applying a series of earthquakes to a board can never produce a previous board state.

We can evaluate $\Phi(GE(f)) - \Phi(f),$ 
\[ \Phi(GE(f)) - \Phi(f) = \sum_{ (x,y)} (GE(f(x,y)) -f(x,y))(|x|+|y|)^2 \]

Let $A$ be the set of locations $(x,y)$ with $f(x,y) \geq 4$, which is non-empty by assumption. We call this the set of active squares. Then we know that $GE(f(x,y)) -f(x,y) \neq 0$ if $(x,y) \in A \cup \bigcup_{(x,y) \in A} n(x,y)$. That is the contribution from square $(x,y)$ to the sum is nonzero only if $(x,y)$ is an active square or it neighbors an active square.

For each $(x,y) \in A$ we know that four chips are removed from the square and one is placed on each of its four neighbors. Thus we have that 
\[ \Phi(GE(f)) - \Phi(f)= \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \]
That is

Now since at most two of the neighbors of $(x,y)$ are a taxi cab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxi cab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2,$ we have that 
\[\sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
 \[ = \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} = \sum_{(x,y) \in A} 4> 0 \] 
\end{proof}

\begin{lemma}
If $LE(f, (x_1, y_1)) \neq f$, then $LE^n(f, \{(x_n,y_n)\}) \neq f$ for all $n > 0$ and all sequences $\{(x_n, y_n)\}$.
\end{lemma}
Our proof of this lemma is essentiall the same as the preceeding one. We construct a potential function and show that if a local earthquake acts nontrivially on a board then the value of the potential function increases.

>>>>

\subsection{A Potential Function}
In order to show finite boards do not cycle, we define a potential function mapping boards to nonnegative integers. We show that earthquakes increase the value of the function. We use the potential function to define a measure of how ``spread out" a board is. So we claim that spread always increases.
\begin{lemma}
If $GE(f) \neq f$, then $GE^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
Let $\Phi: \mathcal{B}_F \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
We can think of $\Phi(f)$ as the sum of the values weighted by the ``taxi-cab" distance from the origin squared. This is well-defined since boards are finite. We claim
\[ GE(f) \neq f \Rightarrow \Phi(GE(f)) > \Phi(f) \]
If a global earthquake changes the board, $\Phi$ increases. This implies if $GE(f) \neq f$, then for all $n > 0$, $GE^n(f) \neq f$. In other words, a series of earthquakes to a finite board can never produce a previous board.

We can evaluate $\Phi(GE(f)) - \Phi(f),$ 
\[ \Phi(GE(f)) - \Phi(f) = \sum_{ (x,y)} (GE(f)(x,y) -f(x,y))(|x|+|y|)^2 \]

Let $A$ be the set of locations $(x,y)$ with $f(x,y) \geq 4$. $A$ is non-empty by assumption. We call $A$ the set of active squares. We know $GE(f)(x,y) -f(x,y) \neq 0$ if $(x,y) \in A \cup \bigcup_{(x,y) \in A} n(x,y)$.

For each $(x,y) \in A$, four chips are removed from the square and one is placed on each of its four neighbors. Thus we have that 
\[ \Phi(GE(f)) - \Phi(f)= \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \]

Since at most two of the neighbors of $(x,y)$ are a taxi cab distance $(|x|+|y|-1)^2$ from the origin and at least two of the neighbors are a taxi cab distance $(|x|+|y|+1)^2$ from the origin and $(|x|+|y|+1)^2 > (|x|+|y|-1)^2,$ we have that 
\[\sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
 \[ = \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} = \sum_{(x,y) \in A} 4> 0 \] 
\end{proof}

\begin{lemma}
Suppose $f \in B_F$. If $LE(f, (x_1, y_1)) \neq f$, then $LE^n(f, \{(x_n,y_n)\}) \neq f$ for all $n > 0$ and all sequences $\{(x_n, y_n)\}$.
\end{lemma}
Our proof of this lemma is essentiall the same as the preceeding one. We construct a potential function and show that if a local earthquake acts nontrivially on a board then the value of the potential function increases.
